---
output: html_document
---

# My first simple Rmd file

Don't forget to install knitr package.

```{r, eval=FALSE}
install.packages("knitr")
```

# Introduction

This is an `.Rmd` document.  Maybe you have heard of `Sweave`?
Well, `knitr` is like modern `Sweave`.

* can work with markdown files (simpler) and LaTeX.
* more flexible re: graphics
* caching
* fancy look and feel.

## A simple example

Let's get R to do some simple maths.

```{r}
x <- 1:3
y <- 4:6
outer(x, y)
```
As well as having chunks in separate paragaphs, you can have some
inline computation.  For example, the mean of x is `r mean(x)` and
the sum of x and y is `r x+y`.  Chunks can be named and then
referred to later.

## Plotting is easy too

```{r plot-chunk}
x <- seq(from=0, to=2*pi, length=1000)
y <- cos(3*x)
plot(x, y, type='l', col='blue')
```

## Tables 

See also the [xtable](http://cran.r-project.org/web/packages/xtable/)
package.

```{r,results='asis'}
#kable(head(iris[,1:3]), format='html')
```

# Okay, what do you need to know?


1. What is markdown syntax? Orignally
   [here](http://daringfireball.net/projects/markdown), promoted for
   `R` by [Rstudio](http://rmarkdown.rstudio.com/).

2. What can `knitr` do? [Knitr home page](http://yihui.name/knitr/)

